A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-14 model.tex
more setup

\section{Single fluid description of dusty gas}\label{setup} 
We model an accretion disk as a mixture of gas with dust treated as a 
pressureless fluid. We denote their density and
velocity field as $(\rhog,\bm{v}_\mathrm{g})$ and
$(\rhod,\bm{v}_\mathrm{d})$, respectively. The two fluids 
interact via a drag force parameterized by a stopping time $\tstop$
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\left(\rhod+\rhog\right)}\frac{\left(\bm{v}_\mathrm{d} -
    \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume. Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rhog/\left(\rhog+\rhod\right)$ defined in previous studies
\citep[e.g.][]{youdin05a,jacquet11}.

@@

%In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
In this limit of strong dust-gas drag, \cite{laibe14} give a single
fluid description of the mixture (their Eq. 84 --- 85 and Eq. 87):

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}, 
\end{align}
and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total mixture density.

In Eq. \ref{dusteq}, 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. Note that Eq. \ref{dusteq} is
equivalent to Eq. 46 of \cite{jacquet11}.

@@

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
To complete the system of equations we need to specify a gas equation
of state. We start with the 
energy equation for an ideal gas of adiabatic index $\gamma$: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  + \mathcal{H} - \Lambda 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
where $\mathcal{H} = 
(\gamma-1)\tepsilon|\rhog\left(\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}\right)|^2$
accounts for heating due to dust-gas drag \citep{laibe14}; and 
\begin{align}
  \Lambda = \frac{\rho \mathcal{R}}{\mu}\frac{T
  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
represents radiative cooling on a characteristic timescale
$\tcool$ that relaxes the gas back to a prescribed temperature profile  
$T_\mathrm{ref}$, where $\mathcal{R},\, \mu$ are  the gas constant and
mean molecular weight of the gas, respectively.

>>>>

\section{Single fluid description of dusty gas}\label{setup} 
We model an accretion disk as a mixture of gas with dust treated as a 
pressureless fluid. We denote their density and
velocity field as $(\rhog,\bm{v}_\mathrm{g})$ and
$(\rhod,\bm{v}_\mathrm{d})$, respectively. The mixture has a single 
temperature $T$, and the mixture pressure $P$ arise solely from the 
gas component.

The two fluids interact via a drag force parameterized by a stopping
time $\tstop$ 
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\left(\rhod+\rhog\right)}\frac{\left(\bm{v}_\mathrm{d} -
    \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume. Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rhog/\left(\rhog+\rhod\right)$ defined in previous studies
\citep[e.g.][]{youdin05a,jacquet11}.

@@

%In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
In this limit of strong dust-gas drag, \cite{laibe14} give a single
fluid description of the mixture: 
\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda, \label{tempeq} 
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}, 
\end{align}
and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total mixture density. For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

In Eq. \ref{dusteq}, 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. Note that Eq. \ref{dusteq} is 
equivalent to Eq. 46 of \cite{jacquet11}.

For the mixture's temperature evolution, Eq. \ref{tempeq}, $\gamma$ is
the adiabatic index; $\mathcal{H}_\mathrm{drag}$ represents heating
due to dust-gas drag and $\mathcal{H}_\mathrm{eff}$ is an effective
energy source arising from transforming from the two-fluid to
one-fluid description \citep[see][ for details]{laibe14}. We also
include a simple model of radiative cooling,
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

@@

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
For later comparison it is convenient to re-write the temperature
evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  + \mathcal{H} 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
where $\mathcal{H}$ collects together dust-drag heating and radiative
cooling.
